---
title: "RaCoCOS: Range-constrained co-occurrence simulation"
author: "Peter Solymos and Zoltan Feher"
date: "`r Sys.Date()`"
output: pdf_document
---

## Introduction

This is supplementary information for the manuscript
*Range-constrained co-occurrence simulation reveals little
niche partitioning among rock-dwelling Montenegrina land snails (Gastropoda: Clausiliidae)* by Z. Feher, K. Jaksch, M. Szekeres, E. Haring, S. Bamberger,
B. Pall-Gergely, and P. Solymos 
(*Journal of Biogeography*, in press).
The document is a tutorial and reproducible example
of the *Range-constrained co-occurrence simulation (RaCoCOS)*
methodology outlined in the manuscript.

## Required packages

The following packages are required to run the code:

```{r reqire,warning=FALSE,message=FALSE,results='hide'}
if (!require(mefa))
    install.packages("mefa")
library(mefa)
if (!require(pbapply))
    install.packages("pbapply")
library(pbapply)
if (!require(vegan))
    install.packages("vegan")
library(vegan)
if (!require(RColorBrewer))
    install.packages("RColorBrewer")
library(RColorBrewer)
if (!require(rworldmap))
    install.packages("rworldmap")
library(rworldmap)
if (!require(sp))
    install.packages("sp")
library(sp)
```

## Distribution data

Use the `setwd` to set working directory where the `Supplementary_data_1.csv`
file can be found.

```{r setwd,echo=FALSE,results='hide'}
setwd("~/Dropbox/collaborations/Montenegrina_simulation/demo/")
```

```{r data}
d <- read.csv("Supplementary_data_1.csv")
## sample identifiers and coordinates
x <- d[,c("ID","code","habitat_type","Latitude","Longitude")]
rownames(x) <- d$ID
summary(x)
## detections
y <- d[,!(colnames(d) %in% colnames(x))]
rownames(y) <- d$ID
y <- as.matrix(y)
y[y > 0] <- 1
str(y)
```

## Definition of taxon ranges

Calculate great circle distance from latitude/longidtude in km's:

```{r gd}
xy <- as.matrix(x[,c("Longitude", "Latitude")])
gd <- spDists(x=xy, longlat=TRUE)
str(gd)
hist(gd)
```


The function `dist_fun` calculates the distance weights
based on the distances (`d`) and the `k` and `d0` arguments:

```{r df}
dist_fun <- function(d, k=5, d0=30) {
  1 / (1 + exp(k * (log(d) - log(d0))))
}
plot(0:100, dist_fun(0:100), type="l")
w <- dist_fun(gd, k=5, d0=30)
str(w)
```

Next we calculate the OP matrix.
The `fun_ij` function calculates the OP for
site `i` and taxon `j`, the `y` arument is the site x taxa table
(the function can be rused for simulated matrices as well).
The `fun_j` function calculates the OP for spacies `j`
at all sites. The `OP_fun` calculates all the OP values
based on an input matrix `y` given the `w` weight matrix
based on the distances and the distance decay function:

```{r op}
fun_ij <- function(i, j, y) {
    sum(w[which(y[,j] > 0), i]) / sum(w[, i])
}
fun_j <- function(j, y) {
    sapply(1:nrow(y), fun_ij, j=j, y=y)
}
OP_fun <- function(y) {
    pbsapply(1:ncol(y), fun_j, y=y)
}
OP <- OP_fun(y)
dimnames(OP) <- dimnames(y)
str(OP)
```

The following function puts the detections or distance weighted
probabilities from the OP matrix onto the map:

```{r map}
plot_distr <-
function(j, type = c("01", "p"), ...)
{
    type <- match.arg(type)
    plot(getMap(resolution = "low"),
         xlim = c(14, 28), ylim = c(36, 47), asp = 1, ...)
    Col <- brewer.pal(9, "Blues")[4:9]
    if (type == "p") {
        z <- cut(OP[,j], breaks=length(Col))
        points(xy[order(z),], pch=19, col = Col[z[order(z)]], cex = .6)
    } else {
        points(xy[y[,j]==0,], pch=".", col=Col[1], cex=1)
        points(xy[y[,j]==1,], pch=19, col=rev(Col)[1], cex=0.6)
    }
    invisible(NULL)
}
plot_distr(32, "01", main="detections")
plot_distr(32, "p", main="distance weighted probabilities")
```

Uncorrected, soft, and hard correction:

```{r corrections}
rnd_fun <- function(y, OP, type="uncorrected") {
    type <- match.arg(type, c("uncorrected", "hard", "soft"))
    yout <- y
    yout[] <- 0
    if (type=="uncorrected") {
        n <- nrow(y)
        for (i in 1:ncol(yout)) {
            yout[sample.int(n, sum(y[,i] > 0), replace=FALSE,
                prob=OP[,i]),i] <- 1
        }
    } else {
        wfun_hard <- function(y) {
            rsum <- rowSums(y)
            sample(rsum)
        }
        wfun_soft <- function(y) {
          rsum <- rowSums(y)+1
          sample(rsum)
        }
        if (type == "hard")
            w <- wfun_hard(y)
        if (type == "soft")
            w <- wfun_soft(y)
        wOP <- w*OP
        yout[sample.int(prod(dim(y)), sum(y > 0),
            prob=wOP)] <- 1
    }
    yout
}
str(rnd_fun(y, OP, "uncorrected"))
str(rnd_fun(y, OP, "hard"))
str(rnd_fun(y, OP, "soft"))
```

## Co-occurrecnce simulations

Now we can calculate co-occurrences. The `coocc_fun` calculates
joint occurrences for species pairs (lower traingle of taxa x taxa matrix).

```{r cooccfun}
coocc_fun <- function(y) {
    co <- t(y) %*% y
    as.dist(co)
}
```

We set number of simulations, and set correction type
(change these settings as desired):

```{r settings}
B <- 99
TYPE <- "uncorrected"
```

We use the summloden infrastructure in the **vegan** package:

```{r}
f <- function(x, n, ...)
    array(replicate(n, rnd_fun(x, OP, type=TYPE)), c(dim(x), n))
(cs <- commsim("dist_prob", fun=f, binary=TRUE,
               isSeq=FALSE, mode="integer"))
(nm <- nullmodel(y, cs))
os <- oecosimu(nm, coocc_fun, nsimul=B)
str(os)
```

Now we list species co-occurrences:

```{r}
str(coocc <- coocc_fun(y))
stack_dist <-
    function (x, dim.names = FALSE, ...)
    {
        id <- as.matrix(x)
        id[lower.tri(id)] <- 1
        id[upper.tri(id)] <- 0
        diag(id) <- 0
        rm <- row(id)
        cm <- col(id)
        rm <- array(rm)[array(id) == 1]
        cm <- array(cm)[array(id) == 1]
        d <- as.vector(x)
        attr(d, "call") <- attr(x, "call")
        attr(d, "method") <- attr(x, "method")
        out <- data.frame(row=rm, col=cm, dist=d)
        if (dim.names) {
            out$row <- as.factor(out$row)
            out$col <- as.factor(out$col)
            levels(out$row) <- rownames(id)[-1]
            levels(out$col) <- colnames(id)[-ncol(id)]
        }
        out
    }
sc <- stack_dist(coocc, dim.names = TRUE)
sc <- sc[order(sc$dist, decreasing=TRUE),]
head(sc)
```

The `d2tab` function makes a sorted table of the distances.
Let's inspect the nullmodel object and plot the results:

```{r}
d2tab <- function(d) {
    if (!inherits(d, "dist"))
        d <- vec2dist(d, attr(coocc, "Size"))
    tab <- stack(d, dim.names=TRUE)
    tab[order(tab$dist, decreasing=TRUE),]
}
## observed co-occurrences
hist(os$statistic)
head(d2tab(os$statistic))
## simulated co-occurrences, average
hist(os$oecosimu$means)
## standardized effect size
hist(os$oecosimu$z)
head(d2tab(os$oecosimu$z))
## two-sided p-values
hist(os$oecosimu$pval)

col <- brewer.pal(9, "RdYlBu")
plot(os$oecosimu$means, os$statistic,
     col=col[cut(os$oecosimu$z, 9)],
     pch=19, cex=0.5)
abline(0,1,col=1)
```

## Session info

```{r session-info}
sessionInfo()
```
